---
title: "Comparison of characteristic features across count data sets"
author: ""
date: ""
output: 
  html_document:
    toc: true
    toc_float: true
    code_folding: show
  pdf_document:
    toc: true
always_allow_html: yes
references:
- id: Love2014DESeq2
  title: Moderated estimation of fold change and dispersion for RNA-seq data with DESeq2
  author:
  - family: Love
    given: Michael I
  - family: Huber
    given: Wolfgang
  - family: Anders
    given: Simon
  container-title: Genome Biology
  volume: 15
  page: 550
  type: article-journal
  URL: https://genomebiology.biomedcentral.com/articles/10.1186/s13059-014-0550-8
  issued:
    year: 2014
- id: Robinson2010edgeR
  title: edgeR-a Bioconductor package for differential expression analysis of digital gene expression data
  author:
  - family: Robinson
    given: Mark D
  - family: McCarthy
    given: Davis J
  - family: Smyth
    given: Gordon K
  container-title: Bioinformatics
  volume: 26
  page: 139-140
  type: article-journal
  URL: https://academic.oup.com/bioinformatics/article-lookup/doi/10.1093/bioinformatics/btp616
  issued:
    year: 2010
- id: Robinson2010TMM
  title: A scaling normalization method for differential expression analysis of RNA-seq data
  author:
  - family: Robinson
    given: Mark D
  - family: Oshlack
    given: Alicia
  container-title: Genome Biology
  volume: 11
  page: R25
  type: article-journal
  URL: https://genomebiology.biomedcentral.com/articles/10.1186/gb-2010-11-3-r25
  issued:
    year: 2010
- id: WaldWolfowitz1940
  title: On a test whether two samples are from the same population
  author:
  - family: Wald
    given: Abraham
  - family: Wolfowitz
    given: Jacob
  container-title: The Annals of Mathematical Statistics
  volume: 11
  page: 147-162
  type: article-journal
  URL: https://projecteuclid.org/euclid.aoms/1177731909
  issued:
    year: 1940
- id: Kolmogorov1933
  title: Sulla determinazione empirica di una legge di distribuzione
  author:
  - family: Kolmogorov
    given: Andrey
  container-title: Giornale dell'Istituto Italiano degli Attuari
  volume: 4
  page: 83-91
  type: article-journal
  issued:
    year: 1933
- id: Smirnov1948
  title: Table for estimating the goodness of fit of empirical distributions
  author:
  - family: Smirnov
    given: Nikolai Vasilyevich
  container-title: Annals of Mathematical Statistics
  volume: 19
  page: 279-281
  type: article-journal
  URL: https://projecteuclid.org/euclid.aoms/1177730256
  issued:
    year: 1948
- id: Rousseeuw1987silhouette
  title: Silhouettes a graphical aid to the interpretation and validation of cluster analysis
  author:
  - family: Rousseeuw
    given: Peter J
  container-title: Journal of Computational and Applied Mathematics
  volumn: 20
  page: 53-65
  type: article-journal
  URL: http://www.sciencedirect.com/science/article/pii/0377042787901257
  issued:
    year: 1987
- id: Chen2014Dispersion
  title: Differential expression analysis of complex RNA-seq experiments using edgeR
  author:
  - family: Chen
    given: Yunshun
  - family: Lun
    given: Aaron TL
  - family: Smyth
    given: Gordon K
  container-title: In Statistical Analysis of Next Generation Sequence Data. Somnath Datta and Daniel S Nettleton (eds), Springer, New York
  URL: https://link.springer.com/chapter/10.1007%2F978-3-319-07212-8_3
  issued:
    year: 2014
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = showCode, warning = FALSE, message = FALSE)
knitr::opts_knit$set(progress = FALSE, verbose = FALSE)
```

```{r preparation, message = FALSE}
## Get the number of data sets
nDatasets <- length(ddsList)

## Determine the number of rows and columns in facetted plots 
## (first element = columns, second element = rows)
colRow <- grDevices::n2mfrow(length(ddsList))

## Set height/width of each panel in multi-panel plots
panelSize <- 4

## Initialize plot list
plots <- list()

## Define plot theme
thm <- theme_bw() + 
  theme(axis.text = element_text(size = 15),
        axis.title = element_text(size = 14),
        strip.text = element_text(size = 15))

## ----------------------- Calculate dispersions ----------------------- ##
if (!quiet) {
  message("Calculating dispersions. This can take some time, please be patient.")
}
if (is.finite(maxNForDisp)) {
  msg <- paste0("If there are more than ", maxNForDisp, 
                " samples in a data set, the dispersions are calculated using ",
                maxNForDisp, " randomly selected samples.")
} else {
  msg <- ""
}
obj <- calculateDispersionsddsList(ddsList = ddsList, 
                                   maxNForDisp = maxNForDisp)

## --------------- Calculate between-sample correlations --------------- ##
if (!quiet) {
  message("Calculating sample correlations.")
}
sampleCorrDF <- calculateSampleCorrs(ddsList = obj, 
                                     maxNForCorr = maxNForCorr)

## -------------- Calculate between-feature correlations --------------- ##
if (!quiet) {
  message("Calculating feature correlations.")
}
featureCorrDF <- calculateFeatureCorrs(ddsList = obj, 
                                       maxNForCorr = maxNForCorr)
  
## ----------------- Summarize sample characteristics ------------------ ##
if (!quiet) {
  message("Summarizing calculated values.")
}
sampleDF <- lapply(obj, function(x) {
  data.frame(
    Libsize = colSums(x$dge$counts),
    Fraczero = colMeans(x$dge$counts == 0),
    TMM = x$dge$samples$norm.factors
  ) %>% dplyr::mutate(EffLibsize = Libsize * TMM)
})
ns <- sapply(sampleDF, nrow)
sampleDF <- do.call(rbind, sampleDF) %>% 
  dplyr::mutate(dataset = rep(names(sampleDF), ns))

## ---------------- Summarize feature characteristics ------------------ ##
featureDF <- lapply(obj, function(x) {
  data.frame(
    Tagwise = sqrt(x$dge$tagwise.dispersion),
    Common = sqrt(x$dge$common.dispersion),
    Trend = sqrt(x$dge$trended.dispersion),
    AveLogCPM = x$dge$AveLogCPM,
    AveLogCPMDisp = x$dge$AveLogCPMDisp, 
    average_log2_cpm = apply(edgeR::cpm(x$dge, prior.count = 2, log = TRUE), 1, mean), 
    variance_log2_cpm = apply(edgeR::cpm(x$dge, prior.count = 2, log = TRUE), 1, var),
    Fraczero = rowMeans(x$dge$counts == 0),
    dispGeneEst = rowData(x$dds)$dispGeneEst,
    dispFit = rowData(x$dds)$dispFit,
    dispFinal = rowData(x$dds)$dispersion,
    baseMeanDisp = rowData(x$dds)$baseMeanDisp,
    baseMean = rowData(x$dds)$baseMean
  )
})
ns <- sapply(featureDF, nrow)
featureDF <- do.call(rbind, featureDF) %>% 
  dplyr::mutate(dataset = rep(names(featureDF), ns))

## ----------------- Summarize data set characteristics ---------------- ##
datasetDF <- do.call(rbind, lapply(obj, function(x) {
  data.frame(
    prior_df = paste0("prior.df = ", round(x$dge$prior.df, 2)),
    nVars = nrow(x$dge$counts),
    nSamples = ncol(x$dge$counts)
  )
})) %>% 
  dplyr::mutate(AveLogCPMDisp = 0.8 * max(featureDF$AveLogCPMDisp)) %>%
  dplyr::mutate(Tagwise = 0.9 * max(featureDF$Tagwise)) %>%
  dplyr::mutate(dataset = names(obj))
```

## Description

`r I(description)`

#### Design formulae

The following model formulae were used in the dispersion calculations for the 
different data sets. Note that if a count matrix or data frame was provided in
place of a *DESeqDataSet* for some data set, the corresponding design formula is
set to `~1`, thus assuming that all samples are to be treated as replicates.

```{r, comment = NA}
## Print design information
for (ds in names(ddsList)) {
  cat(ds, ": ", as.character(DESeq2::design(ddsList[[ds]])), "\n")
}
```

## Data set dimensions {.tabset .tabset-pills}

```{r, include = FALSE}
if (!quiet) {
  message("Processing data set dimensions.")
}
```

These bar plots show the number of samples (columns) and features (rows) in
each data set.

### Number of samples (columns)

```{r nSamples, fig.width = 10, fig.height = 7}
plots[["nSamples"]] <- 
  ggplot(datasetDF, aes(x = dataset, y = nSamples, fill = dataset)) + 
  geom_bar(stat = "identity", alpha = 0.5) + 
  xlab("") + ylab("Number of samples (columns)") + 
  thm + theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))
plots[["nSamples"]]
```

### Number of features (rows)

```{r nVariables, fig.width = 10, fig.height = 7}
plots[["nVariables"]] <- 
  ggplot(datasetDF, aes(x = dataset, y = nVars, fill = dataset)) + 
  geom_bar(stat = "identity", alpha = 0.5) + 
  xlab("") + ylab("Number of features (rows)") + 
  thm + theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))
plots[["nVariables"]]
```

## Dispersion/BCV plots {.tabset .tabset-pills}

```{r, include = FALSE}
if (!quiet) {
  message("Processing dispersion estimates.")
}
```

Disperson/BCV plots show the association between the average abundance and the
dispersion or "biological coefficient of variation" (sqrt(dispersion)), as
calculated by 
[`edgeR`](https://bioconductor.org/packages/release/bioc/html/edgeR.html) 
[@Robinson2010edgeR] and 
[`DESeq2`](http://bioconductor.org/packages/release/bioc/html/DESeq2.html) 
[@Love2014DESeq2]. In the `edgeR` plot, the estimate of the prior degrees of 
freedom is indicated. `r I(msg)`

### edgeR

The black dots represent the tagwise dispersion estimates, the red line the 
common dispersion and the blue curve represents the trended dispersion 
estimates. For further information about the dispersion estimation in `edgeR`,
see @Chen2014Dispersion.

```{r BCVedgeR, fig.width = panelSize * colRow[1], fig.height = panelSize * colRow[2]}
plots[["BCVedgeR"]] <- 
  ggplot(featureDF %>% dplyr::arrange(AveLogCPMDisp), 
         aes(x = AveLogCPMDisp, y = Tagwise)) + 
  geom_point(size = 0.25, alpha = 0.5) + 
  facet_wrap(~dataset, nrow = colRow[2]) + 
  geom_line(aes(y = Trend), color = "blue", size = 1.5) + 
  geom_line(aes(y = Common), color = "red", size = 1.5) +
  geom_text(data = datasetDF, aes(label = prior_df)) + 
  xlab("Average log CPM") + ylab("Biological coefficient of variation") + 
  thm
plots[["BCVedgeR"]]
```

### Pairwise comparisons - edgeR

`r I(defineTableDesc(calculateStatistics = calculateStatistics, subsampleSize = subsampleSize, kfrac = kfrac, kmin = kmin, obstype = "feature", aspect = "average log CPM and tagwise dispersion", minvalue = 0, maxvalue = 1, permutationPvalues = permutationPvalues, nPermutations = nPermutations, nDatasets = nDatasets)$tabledesc2d)`


```{r BCVedgeRTests}
if (calculateStatistics) {
  makeDF(df = featureDF, column = c("AveLogCPMDisp", "Tagwise"), 
         permutationPvalues = permutationPvalues, nPermutations = nPermutations, 
         subsampleSize = subsampleSize, kmin = kmin, kfrac = kfrac)
}
```

### DESeq2

The black dots are the gene-wise dispersion estimates, the red curve the fitted 
mean-dispersion relationship and the blue circles represent the final dispersion
estimates.For further information about the dispersion estimation in `DESeq2`,
see @Love2014DESeq2.

```{r dispersionDESeq2, fig.width = panelSize * colRow[1], fig.height = panelSize * colRow[2]}
plots[["dispersionDESeq2"]] <- 
  ggplot(featureDF %>% dplyr::arrange(baseMeanDisp), 
         aes(x = baseMeanDisp, y = dispGeneEst)) + 
  geom_point(size = 0.25, alpha = 0.5) + 
  facet_wrap(~dataset, nrow = colRow[2]) + scale_x_log10() + scale_y_log10() +  
  geom_point(aes(y = dispFinal), color = "lightblue", shape = 21) + 
  geom_line(aes(y = dispFit), color = "red", size = 1.5) + 
  xlab("Base mean") + ylab("Dispersion") + 
  thm
plots[["dispersionDESeq2"]]
```

### Pairwise comparisons - DESeq2

`r I(defineTableDesc(calculateStatistics = calculateStatistics, subsampleSize = subsampleSize, kfrac = kfrac, kmin = kmin, obstype = "feature", aspect = "base mean and genewise dispersion", minvalue = 0, maxvalue = 1, permutationPvalues = permutationPvalues, nPermutations = nPermutations, nDatasets = nDatasets)$tabledesc2d)`

```{r dispersionDESeq2Tests}
if (calculateStatistics) {
  makeDF(df = featureDF, column = c("baseMeanDisp", "dispGeneEst"), 
         permutationPvalues = permutationPvalues, nPermutations = nPermutations, 
         subsampleSize = subsampleSize, kmin = kmin, kfrac = kfrac)
}
```

## Mean-variance plots {.tabset .tabset-pills}

```{r, include = FALSE}
if (!quiet) {
  message("Processing mean-variance relationship.")
}
```

These scatter plots show the relation between the empirical mean and variance of
the features. The difference between these mean-variance plots and the 
mean-dispersion plots above is that the plots in this section do not take the 
information about the experimental design and sample grouping into account, but 
simply display the mean and variance of log2(CPM) estimates across all samples, 
calculated using the `cpm` function from 
[`edgeR`](https://bioconductor.org/packages/release/bioc/html/edgeR.html) 
[@Robinson2010edgeR], with a prior count of 2.

### Separate scatter plots

```{r meanVarSepScatter, fig.width = panelSize * colRow[1], fig.height = panelSize * colRow[2]}
plots[["meanVarSepScatter"]] <- 
  ggplot(featureDF, aes(x = average_log2_cpm, y = variance_log2_cpm)) + 
  geom_point(size = 0.75, alpha = 0.5) + 
  facet_wrap(~dataset, nrow = colRow[2]) + 
  xlab("Mean of log2(CPM)") + ylab("Variance of log2(CPM)") + 
  thm
plots[["meanVarSepScatter"]]
```

### Overlaid scatter plots

```{r meanVarOverlaidScatter}
plots[["meanVarOverlaidScatter"]] <- 
  ggplot(featureDF, aes(x = average_log2_cpm, y = variance_log2_cpm, color = dataset)) + 
  geom_point(size = 0.75, alpha = 0.5) +
  xlab("Mean of log2(CPM)") + ylab("Variance of log2(CPM)") + 
  thm
plots[["meanVarOverlaidScatter"]]
```

### Pairwise comparisons

`r I(defineTableDesc(calculateStatistics = calculateStatistics, subsampleSize = subsampleSize, kfrac = kfrac, kmin = kmin, obstype = "feature", aspect = "average and variance of the log CPM", minvalue = 0, maxvalue = 1, permutationPvalues = permutationPvalues, nPermutations = nPermutations, nDatasets = nDatasets)$tabledesc2d)` 

```{r meanVarTests}
if (calculateStatistics) {
  makeDF(df = featureDF, column = c("average_log2_cpm", "variance_log2_cpm"), 
         permutationPvalues = permutationPvalues, nPermutations = nPermutations, 
         subsampleSize = subsampleSize, kmin = kmin, kfrac = kfrac)
}
```

## Library sizes {.tabset .tabset-pills}

```{r, include = FALSE}
if (!quiet) {
  message("Processing library sizes.")
}
```

These plots illustrate the distribution of the total read count per sample,
i.e., the column sums of the respective data matrices.

### Separate histograms

```{r libsizeSepHist, fig.width = panelSize * colRow[1], fig.height = panelSize * colRow[2]}
plots[["libsizeSepHist"]] <- 
  ggplot(sampleDF, aes(x = Libsize)) + geom_histogram(bins = 30) + 
  facet_wrap(~dataset, nrow = colRow[2]) +
  xlab("Library size") + thm
plots[["libsizeSepHist"]]
```

### Overlaid histograms

```{r libsizeOverlaidHist}
plots[["libsizeOverlaidHist"]] <- 
  ggplot(sampleDF, aes(x = Libsize, group = dataset, fill = dataset)) + 
  geom_histogram(bins = 30, alpha = 0.5, position = "identity") +
  xlab("Library size") + thm
plots[["libsizeOverlaidHist"]]
```

### Density plots

```{r libsizeDensity}
plots[["libsizeDensity"]] <- 
  ggplot(sampleDF, aes(x = Libsize, group = dataset, color = dataset)) + 
  geom_line(alpha = 0.5, size = 1.5, stat = "density") +
  xlab("Library size") + thm
plots[["libsizeDensity"]]
```

### Density plots (filled)

```{r libsizeDensityFilled}
plots[["libsizeDensityFilled"]] <- 
  ggplot(sampleDF, aes(x = Libsize, group = dataset, fill = dataset)) + 
  geom_density(alpha = 0.5) +
  xlab("Library size") + thm
plots[["libsizeDensityFilled"]]
```

### Box plots

```{r libsizeBox}
plots[["libsizeBox"]] <- 
  ggplot(sampleDF, aes(x = dataset, y = Libsize, color = dataset)) + 
  geom_boxplot(outlier.size = -1, alpha = 0.5) + 
  geom_point(position = position_jitter(width = 0.2), size = 2, alpha = 0.5) + 
  ylab("Library size") + xlab("") + thm + 
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
plots[["libsizeBox"]]
```

### Violin plots

```{r libsizeViolin}
plots[["libsizeViolin"]] <- 
  ggplot(sampleDF, aes(x = dataset, y = Libsize, fill = dataset)) + 
  geom_violin(alpha = 0.5) + 
  ylab("Library size") + xlab("") + thm + 
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
plots[["libsizeViolin"]]
```

### Empirical cumulative distribution function 

```{r libsizeEcdf}
plots[["libsizeEcdf"]] <- 
  ggplot(sampleDF, aes(x = Libsize, group = dataset, color = dataset)) + 
  stat_ecdf(size = 1.5, alpha = 0.5) + ylab("Cumulative proportion") + 
  xlab("Library size") + thm
plots[["libsizeEcdf"]]
```

### Pairwise comparisons

`r I(defineTableDesc(calculateStatistics = calculateStatistics, subsampleSize = subsampleSize, kfrac = kfrac, kmin = kmin, obstype = "sample", aspect = "library size", minvalue = min(sampleDF$Libsize, na.rm = TRUE), maxvalue = max(sampleDF$Libsize, na.rm = TRUE), permutationPvalues = permutationPvalues, nPermutations = nPermutations, nDatasets = nDatasets)$tabledesc)`

```{r libsizeTests}
if (calculateStatistics) {
  makeDF(df = sampleDF, column = "Libsize", 
         permutationPvalues = permutationPvalues, nPermutations = nPermutations, 
         subsampleSize = subsampleSize, kmin = kmin, kfrac = kfrac)
}
```

## TMM normalization factors {.tabset .tabset-pills}

```{r, include = FALSE}
if (!quiet) {
  message("Processing TMM normalization factors.")
}
```

The plots below show the distribution of the TMM normalization factors
[@Robinson2010TMM], intended to adjust for differences in RNA composition, as
calculated by 
[`edgeR`](https://bioconductor.org/packages/release/bioc/html/edgeR.html) 
[@Robinson2010edgeR].

### Separate histograms

```{r tmmSepHist, fig.width = panelSize * colRow[1], fig.height = panelSize * colRow[2]}
plots[["tmmSepHist"]] <- 
  ggplot(sampleDF, aes(x = TMM)) + geom_histogram(bins = 30) + 
  facet_wrap(~dataset, nrow = colRow[2]) +
  xlab("TMM normalization factor") + thm
plots[["tmmSepHist"]]
```

### Overlaid histograms

```{r tmmOverlaidHist}
plots[["tmmOverlaidHist"]] <- 
  ggplot(sampleDF, aes(x = TMM, group = dataset, fill = dataset)) + 
  geom_histogram(bins = 30, alpha = 0.5, position = "identity") +
  xlab("TMM normalization factor") + thm
plots[["tmmOverlaidHist"]]
```

### Density plots

```{r tmmDensity}
plots[["tmmDensity"]] <- 
  ggplot(sampleDF, aes(x = TMM, group = dataset, color = dataset)) + 
  geom_line(alpha = 0.5, size = 1.5, stat = "density") +
  xlab("TMM normalization factor") + thm
plots[["tmmDensity"]]
```

### Density plots (filled) 

```{r tmmDensityFilled}
plots[["tmmDensityFilled"]] <- 
  ggplot(sampleDF, aes(x = TMM, group = dataset, fill = dataset)) + 
  geom_density(alpha = 0.5) +
  xlab("TMM normalization factor") + thm
plots[["tmmDensityFilled"]]
```

### Box plots

```{r tmmBox}
plots[["tmmBox"]] <- 
  ggplot(sampleDF, aes(x = dataset, y = TMM, color = dataset)) + 
  geom_boxplot(outlier.size = -1, alpha = 0.5) + 
  geom_point(position = position_jitter(width = 0.2), size = 2, alpha = 0.5) + 
  ylab("TMM normalization factor") + xlab("") + thm + 
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
plots[["tmmBox"]]
```

### Violin plots

```{r tmmViolin}
plots[["tmmViolin"]] <- 
  ggplot(sampleDF, aes(x = dataset, y = TMM, fill = dataset)) + 
  geom_violin(alpha = 0.5) + 
  ylab("TMM normalization factor") + xlab("") + thm + 
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
plots[["tmmViolin"]]
```

### Empirical cumulative distribution function 

```{r tmmEcdf}
plots[["tmmEcdf"]] <- 
  ggplot(sampleDF, aes(x = TMM, group = dataset, color = dataset)) + 
  stat_ecdf(size = 1.5, alpha = 0.5) + ylab("Cumulative proportion") + 
  xlab("TMM normalization factor") + thm
plots[["tmmEcdf"]]
```

### Pairwise comparisons

`r I(defineTableDesc(calculateStatistics = calculateStatistics, subsampleSize = subsampleSize, kfrac = kfrac, kmin = kmin, obstype = "sample", aspect = "TMM factor", minvalue = min(sampleDF$TMM, na.rm = TRUE), maxvalue = max(sampleDF$TMM, na.rm = TRUE), permutationPvalues = permutationPvalues, nPermutations = nPermutations, nDatasets = nDatasets)$tabledesc)`

```{r tmmTests}
if (calculateStatistics) {
  makeDF(df = sampleDF, column = "TMM", 
         permutationPvalues = permutationPvalues, nPermutations = nPermutations, 
         subsampleSize = subsampleSize, kmin = kmin, kfrac = kfrac)
}
```

## Effective library sizes {.tabset .tabset-pills}

```{r, include = FALSE}
if (!quiet) {
  message("Processing effective library sizes.")
}
```

These plots show the distribution of the "effective library sizes", defined as
the total count per sample multiplied by the corresponding TMM normalization
factor.

### Separate histograms

```{r effLibsizeSepHist, fig.width = panelSize * colRow[1], fig.height = panelSize * colRow[2]}
plots[["effLibsizeSepHist"]] <- 
  ggplot(sampleDF, aes(x = EffLibsize)) + geom_histogram(bins = 30) + 
  facet_wrap(~dataset, nrow = colRow[2]) +
  xlab("Effective library size") + thm
plots[["effLibsizeSepHist"]]
```

### Overlaid histograms

```{r effLibsizeOverlaidHist}
plots[["effLibsizeOverlaidHist"]] <- 
  ggplot(sampleDF, aes(x = EffLibsize, group = dataset, fill = dataset)) + 
  geom_histogram(bins = 30, alpha = 0.5, position = "identity") +
  xlab("Effective library size") + thm
plots[["effLibsizeOverlaidHist"]]
```

### Density plots

```{r effLibsizeDensity}
plots[["effLibsizeDensity"]] <- 
  ggplot(sampleDF, aes(x = EffLibsize, group = dataset, color = dataset)) + 
  geom_line(alpha = 0.5, size = 1.5, stat = "density") +
  xlab("Effective library size") + thm
plots[["effLibsizeDensity"]]
```

### Density plots (filled)

```{r effLibsizeDensityFilled}
plots[["effLibsizeDensityFilled"]] <- 
  ggplot(sampleDF, aes(x = EffLibsize, group = dataset, fill = dataset)) + 
  geom_density(alpha = 0.5) +
  xlab("Effective library size") + thm
plots[["effLibsizeDensityFilled"]]
```

### Box plots

```{r effLibsizeBox}
plots[["effLibsizeBox"]] <- 
  ggplot(sampleDF, aes(x = dataset, y = EffLibsize, color = dataset)) + 
  geom_boxplot(outlier.size = -1, alpha = 0.5) + 
  geom_point(position = position_jitter(width = 0.2), size = 2, alpha = 0.5) + 
  ylab("Effective library size") + xlab("") + thm + 
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
plots[["effLibsizeBox"]]
```

### Violin plots

```{r effLibsizeViolin}
plots[["effLibsizeViolin"]] <- 
  ggplot(sampleDF, aes(x = dataset, y = EffLibsize, fill = dataset)) + 
  geom_violin(alpha = 0.5) + 
  ylab("Effective library size") + xlab("") + thm + 
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
plots[["effLibsizeViolin"]]
```

### Empirical cumulative distribution function 

```{r effLibsizeEcdf}
plots[["effLibsizeEcdf"]] <- 
  ggplot(sampleDF, aes(x = EffLibsize, group = dataset, color = dataset)) + 
  stat_ecdf(size = 1.5, alpha = 0.5) + ylab("Cumulative proportion") + 
  xlab("Effective library size") + thm
plots[["effLibsizeEcdf"]]
```

### Pairwise comparisons

`r I(defineTableDesc(calculateStatistics = calculateStatistics, subsampleSize = subsampleSize, kfrac = kfrac, kmin = kmin, obstype = "sample", aspect = "effective library size", minvalue = min(sampleDF$EffLibsize, na.rm = TRUE), maxvalue = max(sampleDF$EffLibsize, na.rm = TRUE), permutationPvalues = permutationPvalues, nPermutations = nPermutations, nDatasets = nDatasets)$tabledesc)`


```{r effLibsizeTests}
if (calculateStatistics) {
  makeDF(df = sampleDF, column = "EffLibsize", 
         permutationPvalues = permutationPvalues, nPermutations = nPermutations, 
         subsampleSize = subsampleSize, kmin = kmin, kfrac = kfrac)
}
```

## Expression distributions (average log CPM) {.tabset .tabset-pills}

```{r, include = FALSE}
if (!quiet) {
  message("Processing expression distributions.")
}
```

The plots in this section show the distribution of average abundance values for
the features. The abundances are log CPM values calculated by `edgeR`.

### Separate histograms

```{r logCPMSepHist, fig.width = panelSize * colRow[1], fig.height = panelSize * colRow[2]}
plots[["logCPMSepHist"]] <- 
  ggplot(featureDF, aes(x = AveLogCPM)) + geom_histogram(bins = 30) + 
  facet_wrap(~dataset, nrow = colRow[2]) +
  xlab("Average log CPM") + thm
plots[["logCPMSepHist"]]
```

### Overlaid histograms

```{r logCPMOverlaidHist}
plots[["logCPMOverlaidHist"]] <- 
  ggplot(featureDF, aes(x = AveLogCPM, group = dataset, fill = dataset)) + 
  geom_histogram(bins = 30, alpha = 0.5, position = "identity") +
  xlab("Average log CPM") + thm
plots[["logCPMOverlaidHist"]]
```

### Density plots

```{r logCPMDensity}
plots[["logCPMDensity"]] <- 
  ggplot(featureDF, aes(x = AveLogCPM, group = dataset, color = dataset)) + 
  geom_line(alpha = 0.5, size = 1.5, stat = "density") +
  xlab("Average log CPM") + thm
plots[["logCPMDensity"]]
```

### Density plots (filled) 

```{r logCPMDensityFilled}
plots[["logCPMDensityFilled"]] <- 
  ggplot(featureDF, aes(x = AveLogCPM, group = dataset, fill = dataset)) + 
  geom_density(alpha = 0.5) +
  xlab("Average log CPM") + thm
plots[["logCPMDensityFilled"]]
```

### Box plots

```{r logCPMBox}
plots[["logCPMBox"]] <- 
  ggplot(featureDF, aes(x = dataset, y = AveLogCPM, color = dataset)) + 
  geom_boxplot(outlier.size = -1, alpha = 0.5) + 
  geom_point(position = position_jitter(width = 0.2), size = 2, alpha = 0.5) + 
  ylab("Average log CPM") + xlab("") + thm + 
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
plots[["logCPMBox"]]
```

### Violin plots

```{r logCPMViolin}
plots[["logCPMViolin"]] <- 
  ggplot(featureDF, aes(x = dataset, y = AveLogCPM, fill = dataset)) + 
  geom_violin(alpha = 0.5) + 
  ylab("Average log CPM") + xlab("") + thm + 
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
plots[["logCPMViolin"]]
```

### Empirical cumulative distribution function 

```{r logCPMEcdf}
plots[["logCPMEcdf"]] <- 
  ggplot(featureDF, aes(x = AveLogCPM, group = dataset, color = dataset)) + 
  stat_ecdf(size = 1.5, alpha = 0.5) + ylab("Cumulative proportion") + 
  xlab("Average log CPM") + thm
plots[["logCPMEcdf"]]
```

### Pairwise comparisons

`r I(defineTableDesc(calculateStatistics = calculateStatistics, subsampleSize = subsampleSize, kfrac = kfrac, kmin = kmin, obstype = "feature", aspect = "average log CPM", minvalue = min(featureDF$AveLogCPM, na.rm = TRUE), maxvalue = max(featureDF$AveLogCPM, na.rm = TRUE), permutationPvalues = permutationPvalues, nPermutations = nPermutations, nDatasets = nDatasets)$tabledesc)`


```{r logCPMTests}
if (calculateStatistics) {
  makeDF(df = featureDF, column = "AveLogCPM", 
         permutationPvalues = permutationPvalues, nPermutations = nPermutations, 
         subsampleSize = subsampleSize, kmin = kmin, kfrac = kfrac)
}
```

## Fraction zeros per sample {.tabset .tabset-pills}

```{r, include = FALSE}
if (!quiet) {
  message("Processing fraction zeros per sample.")
}
```

These plots show the distribution of the fraction of zeros observed per sample
(column) in the count matrices.

### Separate histograms

```{r fraczeroSampleSepHist, fig.width = panelSize * colRow[1], fig.height = panelSize * colRow[2]}
plots[["fraczeroSampleSepHist"]] <- 
  ggplot(sampleDF, aes(x = Fraczero)) + geom_histogram(bins = 30) + 
  facet_wrap(~dataset, nrow = colRow[2]) +
  xlab("Fraction zeros per sample") + thm
plots[["fraczeroSampleSepHist"]]
```

### Overlaid histograms

```{r fraczeroSampleOverlaidHist}
plots[["fraczeroSampleOverlaidHist"]] <- 
  ggplot(sampleDF, aes(x = Fraczero, group = dataset, fill = dataset)) + 
  geom_histogram(bins = 30, alpha = 0.5, position = "identity") +
  xlab("Fraction zeros per sample") + thm
plots[["fraczeroSampleOverlaidHist"]]
```

### Density plots

```{r fraczeroSampleDensity}
plots[["fraczeroSampleDensity"]] <- 
  ggplot(sampleDF, aes(x = Fraczero, group = dataset, color = dataset)) + 
  geom_line(alpha = 0.5, size = 1.5, stat = "density") +
  xlab("Fraction zeros per sample") + thm
plots[["fraczeroSampleDensity"]]
```

### Density plots (filled) 

```{r fraczeroSampleDensityFilled}
plots[["fraczeroSampleDensityFilled"]] <- 
  ggplot(sampleDF, aes(x = Fraczero, group = dataset, fill = dataset)) + 
  geom_density(alpha = 0.5) +
  xlab("Fraction zeros per sample") + thm
plots[["fraczeroSampleDensityFilled"]]
```

### Box plots

```{r fraczeroSampleBox}
plots[["fraczeroSampleBox"]] <- 
  ggplot(sampleDF, aes(x = dataset, y = Fraczero, color = dataset)) + 
  geom_boxplot(outlier.size = -1, alpha = 0.5) + 
  geom_point(position = position_jitter(width = 0.2), size = 2, alpha = 0.5) + 
  ylab("Fraction zeros per sample") + xlab("") + thm + 
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
plots[["fraczeroSampleBox"]]
```

### Violin plots

```{r fraczeroSampleViolin}
plots[["fraczeroSampleViolin"]] <- 
  ggplot(sampleDF, aes(x = dataset, y = Fraczero, fill = dataset)) + 
  geom_violin(alpha = 0.5) + 
  ylab("Fraction zeros per sample") + xlab("") + thm + 
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
plots[["fraczeroSampleViolin"]]
```

### Empirical cumulative distribution function 

```{r fraczeroSampleEcdf}
plots[["fraczeroSampleEcdf"]] <- 
  ggplot(sampleDF, aes(x = Fraczero, group = dataset, color = dataset)) + 
  stat_ecdf(size = 1.5, alpha = 0.5) + ylab("Cumulative proportion") + 
  xlab("Fraction zeros per sample") + thm
plots[["fraczeroSampleEcdf"]]
```

### Pairwise comparisons

`r I(defineTableDesc(calculateStatistics = calculateStatistics, subsampleSize = subsampleSize, kfrac = kfrac, kmin = kmin, obstype = "sample", aspect = "fraction zeros", minvalue = min(sampleDF$Fraczero, na.rm = TRUE), maxvalue = max(sampleDF$Fraczero, na.rm = TRUE), permutationPvalues = permutationPvalues, nPermutations = nPermutations, nDatasets = nDatasets)$tabledesc)`
 

```{r fraczeroSampleTests}
if (calculateStatistics) {
  makeDF(df = sampleDF, column = "Fraczero", 
         permutationPvalues = permutationPvalues, nPermutations = nPermutations, 
         subsampleSize = subsampleSize, kmin = kmin, kfrac = kfrac)
}
```

## Fraction zeros per feature {.tabset .tabset-pills}

```{r, include = FALSE}
if (!quiet) {
  message("Processing fraction zeros per feature.")
}
```

These plots illustrate the distribution of the fraction of zeros observed per
feature (row) in the count matrices.

### Separate histograms

```{r fraczeroFeatureSepHist, fig.width = panelSize * colRow[1], fig.height = panelSize * colRow[2]}
plots[["fraczeroFeatureSepHist"]] <- 
  ggplot(featureDF, aes(x = Fraczero)) + geom_histogram(bins = 30) + 
  facet_wrap(~dataset, nrow = colRow[2]) +
  xlab("Fraction zeros per feature") + thm
plots[["fraczeroFeatureSepHist"]]
```

### Overlaid histograms

```{r fraczeroFeatureOverlaidHist}
plots[["fraczeroFeatureOverlaidHist"]] <- 
  ggplot(featureDF, aes(x = Fraczero, group = dataset, fill = dataset)) + 
  geom_histogram(bins = 30, alpha = 0.5, position = "identity") +
  xlab("Fraction zeros per feature") + thm
plots[["fraczeroFeatureOverlaidHist"]]
```

### Density plots

```{r fraczeroFeatureDensity}
plots[["fraczeroFeatureDensity"]] <- 
  ggplot(featureDF, aes(x = Fraczero, group = dataset, color = dataset)) + 
  geom_line(alpha = 0.5, size = 1.5, stat = "density") +
  xlab("Fraction zeros per feature") + thm
plots[["fraczeroFeatureDensity"]]
```

### Density plots (filled) 

```{r fraczeroFeatureDensityFilled}
plots[["fraczeroFeatureDensityFilled"]] <- 
  ggplot(featureDF, aes(x = Fraczero, group = dataset, fill = dataset)) + 
  geom_density(alpha = 0.5) +
  xlab("Fraction zeros per feature") + thm
plots[["fraczeroFeatureDensityFilled"]]
```

### Box plots

```{r fraczeroFeatureBox}
plots[["fraczeroFeatureBox"]] <- 
  ggplot(featureDF, aes(x = dataset, y = Fraczero, color = dataset)) + 
  geom_boxplot(outlier.size = -1, alpha = 0.5) + 
  geom_point(position = position_jitter(width = 0.2), size = 2, alpha = 0.5) + 
  ylab("Fraction zeros per feature") + xlab("") + thm + 
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
plots[["fraczeroFeatureBox"]]
```

### Violin plots

```{r fraczeroFeatureViolin}
plots[["fraczeroFeatureViolin"]] <- 
  ggplot(featureDF, aes(x = dataset, y = Fraczero, fill = dataset)) + 
  geom_violin(alpha = 0.5) + 
  ylab("Fraction zeros per feature") + xlab("") + thm + 
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
plots[["fraczeroFeatureViolin"]]
```

### Empirical cumulative distribution function 

```{r fraczeroFeatureEcdf}
plots[["fraczeroFeatureEcdf"]] <- 
  ggplot(featureDF, aes(x = Fraczero, group = dataset, color = dataset)) + 
  stat_ecdf(size = 1.5, alpha = 0.5) + ylab("Cumulative proportion") + 
  xlab("Fraction zeros per feature") + thm
plots[["fraczeroFeatureEcdf"]]
```

### Pairwise comparisons

`r I(defineTableDesc(calculateStatistics = calculateStatistics, subsampleSize = subsampleSize, kfrac = kfrac, kmin = kmin, obstype = "feature", aspect = "fraction zeros", minvalue = min(featureDF$Fraczero, na.rm = TRUE), maxvalue = max(featureDF$Fraczero, na.rm = TRUE), permutationPvalues = permutationPvalues, nPermutations = nPermutations, nDatasets = nDatasets)$tabledesc)`


```{r fraczeroFeatureTests}
if (calculateStatistics) {
  makeDF(df = featureDF, column = "Fraczero", 
         permutationPvalues = permutationPvalues, nPermutations = nPermutations, 
         subsampleSize = subsampleSize, kmin = kmin, kfrac = kfrac)
}
```

## Sample-sample correlations {.tabset .tabset-pills}

```{r, include = FALSE}
if (!quiet) {
  message("Processing sample-sample correlations.")
}
```

The plots below show the distribution of Spearman correlation coefficients for
pairs of samples, calculated from the log(CPM) values obtained via the `cpm`
function from `edgeR`, with a prior.count of 2. If there are more than 
`r maxNForCorr` samples in a data set, the pairwise correlations between 
`r maxNForCorr` randomly selected samples are shown.

### Separate histograms

```{r sampleCorrSepHist, fig.width = panelSize * colRow[1], fig.height = panelSize * colRow[2]}
plots[["sampleCorrSepHist"]] <- 
  ggplot(sampleCorrDF, aes(x = Correlation)) + geom_histogram(bins = 30) + 
  facet_wrap(~dataset, nrow = colRow[2]) +
  xlab("Sample-sample correlation") + thm
plots[["sampleCorrSepHist"]]
```

### Overlaid histograms

```{r sampleCorrOverlaidHist}
plots[["sampleCorrOverlaidHist"]] <- 
  ggplot(sampleCorrDF, aes(x = Correlation, group = dataset, fill = dataset)) + 
  geom_histogram(bins = 30, alpha = 0.5, position = "identity") +
  xlab("Sample-sample correlation") + thm
plots[["sampleCorrOverlaidHist"]]
```

### Density plots

```{r sampleCorrDensity}
plots[["sampleCorrDensity"]] <- 
  ggplot(sampleCorrDF, aes(x = Correlation, group = dataset, color = dataset)) + 
  geom_line(alpha = 0.5, size = 1.5, stat = "density") +
  xlab("Sample-sample correlation") + thm
plots[["sampleCorrDensity"]]
```

### Density plots (filled)

```{r sampleCorrDensityFilled}
plots[["sampleCorrDensityFilled"]] <- 
  ggplot(sampleCorrDF, aes(x = Correlation, group = dataset, fill = dataset)) + 
  geom_density(alpha = 0.5) +
  xlab("Sample-sample correlation") + thm
plots[["sampleCorrDensityFilled"]]
```

### Box plots

```{r sampleCorrBox}
plots[["sampleCorrBox"]] <- 
  ggplot(sampleCorrDF, aes(x = dataset, y = Correlation, color = dataset)) + 
  geom_boxplot(outlier.size = -1, alpha = 0.5) + 
  geom_point(position = position_jitter(width = 0.2), size = 2, alpha = 0.5) + 
  ylab("Sample-sample correlation") + xlab("") + thm + 
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
plots[["sampleCorrBox"]]
```

### Violin plots

```{r sampleCorrViolin}
plots[["sampleCorrViolin"]] <- 
  ggplot(sampleCorrDF, aes(x = dataset, y = Correlation, fill = dataset)) + 
  geom_violin(alpha = 0.5) + 
  ylab("Sample-sample correlation") + xlab("") + thm + 
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
plots[["sampleCorrViolin"]]
```

### Empirical cumulative distribution function 

```{r sampleCorrEcdf}
plots[["sampleCorrEcdf"]] <- 
  ggplot(sampleCorrDF, aes(x = Correlation, group = dataset, color = dataset)) + 
  stat_ecdf(size = 1.5, alpha = 0.5) + ylab("Cumulative proportion") + 
  xlab("Sample-sample correlation") + thm
plots[["sampleCorrEcdf"]]
```

### Pairwise comparisons

`r I(defineTableDesc(calculateStatistics = calculateStatistics, subsampleSize = subsampleSize, kfrac = kfrac, kmin = kmin, obstype = "sample pair", aspect = "Spearman correlation", minvalue = min(sampleCorrDF$Correlation, na.rm = TRUE), maxvalue = max(sampleCorrDF$Correlation, na.rm = TRUE), permutationPvalues = permutationPvalues, nPermutations = nPermutations, nDatasets = nDatasets)$tabledesc)`
 

```{r sampleCorrTests}
if (calculateStatistics) {
  makeDF(df = sampleCorrDF, column = "Correlation", 
         permutationPvalues = permutationPvalues, nPermutations = nPermutations, 
         subsampleSize = subsampleSize, kmin = kmin, kfrac = kfrac)
}
```

## Feature-feature correlations {.tabset .tabset-pills}

```{r, include = FALSE}
if (!quiet) {
  message("Processing feature-feature correlations.")
}
```

These plots illustrate the distribution of Spearman correlation coefficients for
pairs of features, calculated from the log(CPM) values obtained via the `cpm`
function from `edgeR`, with a prior.count of 2. Only non-constant features are
considered, and if there are more than `r maxNForCorr` such features in a
data set, the pairwise correlations between `r maxNForCorr` randomly selected
features are shown.

### Separate histograms

```{r featureCorrSepHist, fig.width = panelSize * colRow[1], fig.height = panelSize * colRow[2]}
plots[["featureCorrSepHist"]] <- 
  ggplot(featureCorrDF, aes(x = Correlation)) + geom_histogram(bins = 30) + 
  facet_wrap(~dataset, nrow = colRow[2]) +
  xlab("Feature-feature correlation") + thm
plots[["featureCorrSepHist"]]
```

### Overlaid histograms

```{r featureCorrOverlaidHist}
plots[["featureCorrOverlaidHist"]] <- 
  ggplot(featureCorrDF, aes(x = Correlation, group = dataset, fill = dataset)) + 
  geom_histogram(bins = 30, alpha = 0.5, position = "identity") +
  xlab("Feature-feature correlation") + thm
plots[["featureCorrOverlaidHist"]]
```

### Density plots

```{r featureCorrDensity}
plots[["featureCorrDensity"]] <- 
  ggplot(featureCorrDF, aes(x = Correlation, group = dataset, color = dataset)) + 
  geom_line(alpha = 0.5, size = 1.5, stat = "density") +
  xlab("Feature-feature correlation") + thm
plots[["featureCorrDensity"]]
```

### Density plots (filled)

```{r featureCorrDensityFilled}
plots[["featureCorrDensityFilled"]] <- 
  ggplot(featureCorrDF, aes(x = Correlation, group = dataset, fill = dataset)) + 
  geom_density(alpha = 0.5) +
  xlab("Feature-feature correlation") + thm
plots[["featureCorrDensityFilled"]]
```

### Box plots

```{r featureCorrBox}
plots[["featureCorrBox"]] <- 
  ggplot(featureCorrDF, aes(x = dataset, y = Correlation, color = dataset)) + 
  geom_boxplot(outlier.size = -1, alpha = 0.5) + 
  geom_point(position = position_jitter(width = 0.2), size = 2, alpha = 0.5) + 
  ylab("Feature-feature correlation") + xlab("") + thm + 
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
plots[["featureCorrBox"]]
```

### Violin plots

```{r featureCorrViolin}
plots[["featureCorrViolin"]] <- 
  ggplot(featureCorrDF, aes(x = dataset, y = Correlation, fill = dataset)) + 
  geom_violin(alpha = 0.5) + 
  ylab("Feature-feature correlation") + xlab("") + thm + 
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
plots[["featureCorrViolin"]]
```

### Empirical cumulative distribution function 

```{r featureCorrEcdf}
plots[["featureCorrEcdf"]] <- 
  ggplot(featureCorrDF, aes(x = Correlation, group = dataset, color = dataset)) + 
  stat_ecdf(size = 1.5, alpha = 0.5) + ylab("Cumulative proportion") + 
  xlab("Feature-feature correlation") + thm
plots[["featureCorrEcdf"]]
```

### Pairwise comparisons

`r I(defineTableDesc(calculateStatistics = calculateStatistics, subsampleSize = subsampleSize, kfrac = kfrac, kmin = kmin, obstype = "feature pair", aspect = "Spearman correlation", minvalue = min(featureCorrDF$Correlation, na.rm = TRUE), maxvalue = max(featureCorrDF$Correlation, na.rm = TRUE), permutationPvalues = permutationPvalues, nPermutations = nPermutations, nDatasets = nDatasets)$tabledesc)`
 

```{r featureCorrTests}
if (calculateStatistics) {
  makeDF(df = featureCorrDF, column = "Correlation", 
         permutationPvalues = permutationPvalues, nPermutations = nPermutations, 
         subsampleSize = subsampleSize, kmin = kmin, kfrac = kfrac)
}
```

## Library size vs fraction zeros {.tabset .tabset-pills}

```{r, include = FALSE}
if (!quiet) {
  message("Processing library size vs fraction zeros relationship.")
}
```

These scatter plots show the association between the total count (column sums)
and the fraction of zeros observed per sample.

### Separate scatter plots

```{r libsizeFraczeroSepScatter, fig.width = panelSize * colRow[1], fig.height = panelSize * colRow[2]}
plots[["libsizeFraczeroSepScatter"]] <- 
  ggplot(sampleDF, aes(x = Libsize, y = Fraczero)) + 
  geom_point(size = 1, alpha = 0.5) + 
  facet_wrap(~dataset, nrow = colRow[2]) + 
  xlab("Library size") + ylab("Fraction zeros") + thm
plots[["libsizeFraczeroSepScatter"]]
```

### Overlaid scatter plots

```{r libsizeFraczeroOverlaidScatter}
plots[["libsizeFraczeroOverlaidScatter"]] <- 
  ggplot(sampleDF, aes(x = Libsize, y = Fraczero, color = dataset)) + 
  geom_point(size = 1, alpha = 0.5) + 
  geom_line(stat = "smooth", method = "loess", size = 1, alpha = 0.5) + 
  xlab("Library size") + ylab("Fraction zeros") + thm
plots[["libsizeFraczeroOverlaidScatter"]]
```

### Pairwise comparisons

`r I(defineTableDesc(calculateStatistics = calculateStatistics, subsampleSize = subsampleSize, kfrac = kfrac, kmin = kmin, obstype = "sample", aspect = "library size and fraction of zeros", minvalue = 0, maxvalue = 1, permutationPvalues = permutationPvalues, nPermutations = nPermutations, nDatasets = nDatasets)$tabledesc2d)` 

```{r libsizeFraczeroTests}
if (calculateStatistics) {
  makeDF(df = sampleDF, column = c("Libsize", "Fraczero"), 
         permutationPvalues = permutationPvalues, nPermutations = nPermutations, 
         subsampleSize = subsampleSize, kmin = kmin, kfrac = kfrac)
}
```

## Mean expression vs fraction zeros {.tabset .tabset-pills}

```{r, include = FALSE}
if (!quiet) {
  message("Processing expression vs fraction zeros relationship.")
}
```

These scatter plots show the association between the average abundance and the 
fraction of zeros observed per feature. The abundance is defined as the
log(CPM) values as calculated by `edgeR`.

### Separate scatter plots

```{r logCPMFraczeroSepScatter, fig.width = panelSize * colRow[1], fig.height = panelSize * colRow[2]}
plots[["logCPMFraczeroSepScatter"]] <- 
  ggplot(featureDF, aes(x = AveLogCPM, y = Fraczero)) + 
  geom_point(size = 0.75, alpha = 0.5) + 
  facet_wrap(~dataset, nrow = colRow[2]) + 
  xlab("Average log CPM") + ylab("Fraction zeros") + thm
plots[["logCPMFraczeroSepScatter"]]
```

### Overlaid scatter plots

```{r logCPMFraczeroOverlaidScatter}
plots[["logCPMFraczeroOverlaidScatter"]] <- 
  ggplot(featureDF, aes(x = AveLogCPM, y = Fraczero, color = dataset)) + 
  geom_point(size = 0.75, alpha = 0.5) + 
  xlab("Average log CPM") + ylab("Fraction zeros") + thm
plots[["logCPMFraczeroOverlaidScatter"]]
```

### Pairwise comparisons

`r I(defineTableDesc(calculateStatistics = calculateStatistics, subsampleSize = subsampleSize, kfrac = kfrac, kmin = kmin, obstype = "feature", aspect = "average log CPM and fraction of zeros", minvalue = 0, maxvalue = 1, permutationPvalues = permutationPvalues, nPermutations = nPermutations, nDatasets = nDatasets)$tabledesc2d)` 

```{r logCPMFraczeroTests}
if (calculateStatistics) {
  makeDF(df = featureDF, column = c("AveLogCPM", "Fraczero"), 
         permutationPvalues = permutationPvalues, nPermutations = nPermutations, 
         subsampleSize = subsampleSize, kmin = kmin, kfrac = kfrac)
}
```


```{r, include = FALSE}
if (savePlots) {
  if (!quiet) {
    message("Saving plots.")
  }
  saveRDS(plots, 
          file = paste0(outputDir, "/",
                        tools::file_path_sans_ext(basename(outputFile)), "_ggplots.rds"))
}
```

## Session info

```{r}
sessionInfo()
```

## References

