---
title: "countsimQC - Comparing characteristic features across count data sets"
author: "Charlotte Soneson"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    fig_caption: yes
vignette: >
  %\VignetteIndexEntry{countsimQC User Guide}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

The `countsimQC` package provides a simple way of comparing characteristic
features across a collection of (RNA-seq) count data sets. An important
application is in situations where a synthetic count data set has been generated
using a real count data set as an underlying source of parameters and 
characteristic features, in which case it is important to verify that the final 
synthetic data in fact does capture the main features of the original data set.
We hope that the reports generated by the package can provide a "standardized"
way to verify the appropriateness of simulated data sets used for evaluation of
computational methods. In this vignette we show how to generate a report from a
collection of two simulated data sets and the original, underlying real data
set. However, the package can be used to compare any collection of count data
sets in terms of their characteristic features, and is thus not limited to this
use case.

First, we load the object containing the three data sets. The object is a named
list, where each element is a \code{DESeqDataSet} object, containing the count
matrix, a sample information data frame and a model formula (necessary to
calculate dispersions).

```{r}
suppressPackageStartupMessages(library(countsimQC))
suppressPackageStartupMessages(library(DESeq2))

data(countsim_example)
countsim_example
```

Next, we generate the report using the `generate_report()` function. 

```{r}
generate_report(dds_list = countsim_example, show_code = FALSE, 
                output_format = "html_document", 
                output_file = "countsim_report.html", output_dir = "./")
```

The report can be generated either as an HTML file (by setting `output_format = 
"html_document"`) or as a pdf file (by setting `output_format =
"pdf_document"`), or both (by setting `output_format = "all"`).

## Session info

```{r}
sessionInfo()
```

